---
title: "Simulation of heterogenous correlation: effects on inference in mixed effects modeling"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulation of heterogeneous correlation: effects on inference in mixed effects modeling}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction; model

In a certain study of methylation patterns in COPD, longitudinal
data on methylation patterns in blood
are collected in a complex design.  For a fraction
of participants, samples from two time points are
assayed on a single chip.  For the other participants,
the samples from the time points are assayed on
different chips.

We will use a very simple setup based on bivariate
Gaussian responses with different means and correlation
coefficients to reflect this design.  We assume
the two groups of donors are each of size $N$, and that
the pairs of observations are obtained at common
times denoted 0 and 1.

Donors in group 1 produce samples according to
$$
Y_i \sim N_2((0,0)^t, 1, \rho_{11}),
$$
using the notation $N_2(\mu, \sigma^2, \rho)$
to denote the bivariate Gaussian distribution with
mean vector $\mu$, variance $\sigma^2$ and correlation coefficient $\rho$.

Donors in group 2 produce samples according to
$$
Y_i \sim N_2((0,b)^t, 1, \rho_{12}).
$$
Both $\rho_{11}$ and $\rho_{12}$ must lie in $[0,1)$.

The parameter of interest is $b$, the difference
in mean response between
groups 1 and 2 arising from the passage of one unit of time.

# Simulation setup

We will assess various modeling approaches via simulation.
We can easily simulate from the models defined above.

We will show that fitting a misspecified model in
which the group-specific intrapair correlations
are assumed common when in fact they are not
leads to a loss of power.

## Code

We'll use MASS to generate multivariate normal
deviates, and nlme to fit mixed effects models.
```{r setup,message=FALSE}
library(MASS)
library(nlme)
```

This function produces the $p$-value for the test of $H_0: b = 0$,
using a random intercept with common variance to capture
longitudinal dependence.

```{r dofun}
simdat1 = function(N=100, b=0, r11=.6, r12=.6) {

# set up correlation matrices
 cm1 = matrix(c(1,r11,r11,1),nc=2)
 cm2 = matrix(c(1,r12,r12,1),nc=2)

# simulate residuals
 S1 = as.numeric(t(mvrnorm(N, c(0,0), Sig=cm1)))
 S2 = as.numeric(t(mvrnorm(N, c(0,0), Sig=cm2)))

# set up ID and time variables, and group indicator
 ids = rep(seq_len(2*N),each=2)
 grp = rep(c(0,1),each=2*N)
 time = rep(c(0,1),2*N)

# define responses: S1 has mean 0, S2 has mean b*time
 dat = data.frame(Y=c(S1,S2+b*time[seq_len(2*N)]), id=ids, time=time, grp=grp)
 dat
}
 
# extract p-value for Ho: b=0
sim1lme = function(dat) {
 mod = lme(Y~time, random=~1|id, data=dat)
 pint = summary(mod)$tTable[2,5]
 list(mod=mod, pint=pint)
}

sim1 = function(N=100, b=0, r11=.6, r12=.6) {
 d = simdat1(N, b, r11, r12)
 sim1lme(d)
}
```

## Some sanity checks

This shows that the code captures the $p$-value of interest.

```{r chk1}
set.seed(123)
f1 = sim1()
summary(f1$mod)
f1$pint
```

When $b$ is fixed at zero, the rejection rate based on $p<0.05$
is as expected.

```{r lksim}
set.seed(123)
library(parallel)
options(mc.cores=6)
```

```{r lkss,cache=TRUE}
nn = unlist(mclapply(seq_len(1500), function(x) sim1()$pint))
mean(nn<0.05)
```

# Power curves for correctly specified and misspecified mixed effects models

For a range of values of $b$, we will plot the rejection
frequencies over 1500 realizations of the model
with $\rho_{12} = $\rho_{11} = 0.6$.

```{r dopow66, cache=TRUE}
bs = seq(0,.5,.05)
pows_66 = lapply(bs,
 function(b) 
   unlist(mclapply(seq_len(1500), function(x) sim1(b=b, r11=.6, r12=.6)$pint))
 )
plot(bs,unlist(sapply(pows_66, function(x) mean(x<0.05))),xlab="b", ylab="power with N=100 per group",
  main = "common rho = .6")
```

Now set $\rho_{12} = 0.2$.

```{r dopow62, cache=TRUE}
pows_62 = lapply(bs,
 function(b) 
   unlist(mclapply(seq_len(1500), function(x) sim1(b=b, r11=.6, r12=.2)$pint))
 )
plot(bs,unlist(sapply(pows_62, function(x) mean(x<0.05))),xlab="b", ylab="power with N=100 per group",
  main = "rho11 = .6, rho12 = .2")
```

Superimposing:

```{r dobo}
p66 = unlist(sapply(pows_66, function(x) mean(x<0.05)))
p62 = unlist(sapply(pows_62, function(x) mean(x<0.05)))
plot(bs, p66, type="l", lty=1)
lines(bs, p62, lty=2)
legend(0,.8, lty=c(1,2), legend=c("common", "heterogeneous"))
```

# Does adding a dummy variable help restore power?

```{r lk1g}
sim2lme = function(dat) {
 mod = lme(Y~time+grp, random=~1|id, data=dat)
 pint = summary(mod)$tTable[2,5]
 list(mod=mod, pint=pint)
}
sim2 = function(N=100, b=0, r11=.6, r12=.6) {
 d = simdat1(N, b, r11, r12)
 sim2lme(d)
}
# type 1 err
nn = unlist(mclapply(seq_len(1500), function(x) sim2()$pint))
print(mean(nn<0.05))
pows_62d = lapply(bs,
 function(b) 
   unlist(mclapply(seq_len(1500), function(x) sim2(b=b, r11=.6, r12=.2)$pint))
 )
p62d = unlist(sapply(pows_62d, function(x) mean(x<0.05)))
plot(bs, p66, type="l", lty=1)
lines(bs, p62, lty=2)
lines(bs, p62d, lty=3)
legend(0,.8, lty=c(1,2,3), legend=c("common", "heterogeneous", "dummy"))
```
